---
title: 'Exercises: Deep Learning with R - Chapter 2'
author: "Guillaume Guénard"
date: '2022-04-12'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data preparation

## Packages, directories, and sources

The following code loads the necessary `R` packages, resources directories, and
sources:

```{r}
library(keras)
library(magrittr)
library(caret)
data_dir <- "../Data"
aux_dir <- "../Auxiliary"
aux_dir %>% file.path("Auxiliary.R") %>% source
```

## Loading the MNIST data set example

This code is for loading the MNIST handwritten digits example data set. To avoid
having to download the data set each time the document is compiled, I have put
the code to download and arrange the data and saves them into an
`if(FALSE){...}` block. The pre-processed data are then simply loaded whenever
the document is compiled.

```{r, fig.width=7, fig.height=5, cache=TRUE}
if(FALSE){
  mnist <- dataset_mnist()
  train_images <- mnist$train$x
  train_labels <- mnist$train$y
  test_images <- mnist$test$x
  test_labels <- mnist$test$y
  rm(mnist)
  save(
    train_images, train_labels, test_images, test_labels,
    file = data_dir %>% file.path("mnist.rda")
  )
} else 
  load(file = data_dir %>% file.path("mnist.rda"))

plot_tile(x=train_images, y=train_labels, n=16L, m=12L)
```

The last code line calls a custom plotting function, defined in the sources, and
that allows one to plot a tile of digit images together with their labels. The
numbers of columns and rows of digits that are displayed are controlled by
arguments `n` and `m`, respectively.

## Reshaping the array

The image data set is a 3-d array (images,width,height) of grayscale values. To
prepare the data for training we convert the 3-d arrays into matrices by
reshaping width and height into a single dimension (i.e., 28x28 images are
flattened into length 784 vectors). Then, we convert the grayscale values from
integers ranging between 0 to 255 into floating point values ranging between 0
and 1. Here, we need function `reticulate::array_reshape()` because the images
are stored using column-major ordering, which is customarily used for storing
data in `R` (see, `help("array_reshape",package="reticulate")` for details).

```{r, cache=TRUE}
train_images %<>% array_reshape(c(nrow(.), 28L*28L))
test_images %<>% array_reshape(c(nrow(.), 28L*28L))
```

Changing the scaling of the values into the 0-1 interval:

```{r, cache=TRUE}
train_images %<>% {./255}
test_images %<>% {./255}
```

Changing the coding of the label to categorical (binary) coding:

```{r, cache=TRUE}
train_labels %<>% to_categorical(10)
test_labels %<>% to_categorical(10)
```

Instead of single vector of integer, we now have a binary matrix:

```{r}
train_labels %>% head
```

# Keras sequential model

Perhaps the simplest type of ANN is a sequential model whereby a single input
layers feeds a chain of hidden layers culminating into a single output layer.
This type of ANN is implemented using function `keras_model_sequential` as
follows:

```{r}
keras_model_sequential() %>%
  layer_dense(
    input_shape = c(28L*28L),
    units = 256,
    activation = 'relu',
    name = "dense_1"
  ) %>%
  layer_dense(
    units = 10,
    activation = 'softmax',
    name = "output"
  ) -> ann_model
```

The resulting model (`ann_model`) has a dense hidden layer with $256$ nodes.
Because it inherits from `keras_model_sequential()` rather than an explicit
upstream layer, function `layer_dense` needs to be provided its input shape
using argument `input_shape`. The $28 \times 28$ images matrices were vectorized
into length $784$ vectors, and thus the input shape is `c(784)`. The output of
the first layer has activation function `'relu'`, which is a simple
defined-by-parts non-linear function returning positive values unchanged and $0$
for the negative values:
$$
RELU(x) = 
\left\{
    \begin{array}{lr}
        x, & \text{if } x > 0\\
        0, & \text{if } x \leq 0
    \end{array}
\right\}
$$
The output of the dense layer is fed directly into a second dense layer with a
softmax output. The softmax function outputs the probability of different
outcomes among the complete set of possible outcomes for a given problem. Here
we have 10 digits, and thus the softmax will output 10 probabilities summing to
$1$ for each image. Each probability is that of the image being associated with
a particular label according to the model. The softmax function is defined as
follows:
$$
\text{softmax}(x_i) = {\text{e}^{x_i} \over \sum_{\forall j} \text{e}^{x_j}}
$$

## Compiling and training the model 

Before being trained, the model first needs to be compiled as follows:

```{r}
ann_model %>%
  compile(
    loss = 'categorical_crossentropy',
    optimizer = optimizer_rmsprop(),
    metrics = c('accuracy')
  )
```

Argument `loss` is a textual variable specifying the loss function to be used
for training the model. Argument `optimizer` specified the optimizer used to
train the model (namely, to perform the gradient descent of the loss function
over the model weights). Argument `metrics` specifies the metrics to estimate
the model success (here: 'accuracy', which means the proportion of the labels
that are correctly called).

The model is then fit to the training data as follows:

```{r, eval=FALSE}
ann_model %>%
  fit(
    train_images,
    train_labels,
    epochs = 60,
    batch_size = 120,
    validation_split = 0.2
  ) -> history
```
```{r, eval=FALSE, echo=FALSE}
ann_model %>%
  save_model_hdf5(
    filepath = "my_model.h5",
    overwrite = TRUE,
    include_optimizer = TRUE
  )
save(history, file = "history.rda")
```
```{r, echo=FALSE}
load_model_hdf5("my_model.h5") -> ann_model
load(file = "history.rda")
```

Its training history is plotted as follows:

```{r, fig.width=7, fig.height=5}
history %>% plot
```

The fitted model (i.e., its structure, weights, and optimizer state) can be
saved to the hdf5 format and then loaded as follows:

```{r, eval=FALSE}
ann_model %>%
  save_model_hdf5(
    filepath = "my_model.h5",
    overwrite = TRUE,
    include_optimizer = TRUE
  )
load_model_hdf5("my_model.h5") -> ann_model
```

Also, its is possible to save the model structure as a JSON string as follows:

```{r}
ann_model %>% model_to_json -> json_string
json_string %>% cat(file="my_model.json")
```

The JSON string will not save the model weights. The latter can be saved as an
hdf5 file as follows:

```{r}
ann_model %>%
  save_model_weights_hdf5(
    filepath = "my_model_weights.h5",
    overwrite = TRUE
  )
```

The model can be loaded from these storage files as follows:

```{r}
model_from_json(json_string) -> model0
model0 %>% load_model_weights_hdf5('my_model_weights.h5')
```

## Evaluating the model's out-of-the-sample accuracy

The model's performance metrics are evaluated from the test images and labels as
follows:

```{r, eval=FALSE}
ann_model %>%
  evaluate(
    test_images,
    test_labels
  ) -> metrics
```
```{r, eval=FALSE, echo=FALSE}
save(metrics, file="metrics.rda")
```
```{r, echo=FALSE}
load(file="metrics.rda")
metrics
```

One obtains the model's predicted test labels as follows:

```{r, eval=FALSE}
ann_model %>%
  predict(test_images) %>%
  apply(1L, function(x) which.max(x) - 1L) %>%
  as.factor -> fitted_labels
```
```{r, eval=FALSE, echo=FALSE}
save(fitted_labels, file="fitted_labels.rda")
```
```{r, echo=FALSE}
load(file="fitted_labels.rda")
```

The confusion matrix is obtained as follows:

```{r, eval=FALSE}
test_labels %>%
  apply(1L, function(x) which.max(x) - 1L) %>%
  as.factor -> reference_labels
fitted_labels %>%
  confusionMatrix(
    reference_labels
  ) -> cmat
cmat
```
```{r, eval=FALSE, echo=FALSE}
save(cmat, file="Confusion_matrix.rda")
```
```{r, echo=FALSE}
load(file="Confusion_matrix.rda")
cmat
```

Some errors are more common than others. For instances, digits 5 was taken to
be a 3, and 7 were taken to be 2 more often than 4 were confused for 1. This is
not surprising from general knowledge.
